---
title: "Copula Modeling for Clinical Trials"
author: "Nathan James"
date: "February 26, 2019"
output: 
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

# for options see: 
# https://bookdown.org/yihui/rmarkdown/ioslides-presentation.html
# http://data-analytics.net/cep/Schedule_files/presentations_demo.html
# https://stackoverflow.com/questions/39035308/how-to-include-plotly-in-r-studio-presentations-rpres

library(plotly)

```

# Introduction

This is an R Markdown presentation. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.

## Background

- Bullet 1
- Bullet 2
- Bullet 3

## Copulas for Joint Modeling

```{r cars, echo = TRUE}
summary(cars)
```

## Applications

```{r pressure}
plot(pressure)
```


# Copula Models

Copula models are super rad

## Notation, Definitions, and Sklar's Theorem

```{r, echo = FALSE, fig.align = 'center', message = FALSE}
plot_ly(z = ~volcano, type = "surface")
```


## Copula Concepts

## Inference

## Copula Regression

## Extensions

# Applications

## Benefit-Risk Analysis

some stuff about benefit-risk

---

more benefit-risk

## Clustered Data

## Discussion and Future Research