
# (APPENDIX) Appendix {-}

# Technical Supplement {#tech-supp}

This appendix has the technical details used to derive copula likelihood function, etc.

* Probability Integral Transformation

Let $Y$ be a continuous random variable with df $F$ and quantile function $F^{-1}$ and define $F(Y)=Z$ then
$P(Z \le z) = P(F(Y) \le z)= P(F^{-1}(F(Y)) \le F^{-1}(z)) = P(Y \le F^{-1}(z))=F(F^{-1}(z)) =z$ which is recognized as the df of a Uniform(0,1) random variable. So $Z \sim Unif(0,1)$

* Formula for C-volume with arbitrary $d$

$\Delta_{(\mathbf{a},\mathbf{b}]}C=\sum_{i \in \{0,1\}^d} (-1)^{\sum_{j=1}^d i_j}C(a_{1}^{i_1}b_{1}^{1-i_1},\ldots,a_{d}^{i_d}b_{d}^{1-i_d})$ where the sum is over the set of $2^d$ vectors of size $d$ with elements 0 or 1. For $d=2$, the set is $\{(i_1=0,i_2=0),(i_1=0,i_2=1),(i_1=1,i_2=0),(i_1=1,i_2=1)\}$ and the C-volume is $\Delta_{(\mathbf{a},\mathbf{b}]}C=C(b_1,b_2)-C(b_1,a_2)-C(a_1,b_2)+C(a_1,a_2)$. For $d=3$ the set is $\{(i_1=0,i_2=0,i_3=0),(i_1=0,i_2=0,i_3=1),(i_1=0,i_2=1,i_3=0),(i_1=0,i_2=1,i_3=1),(i_1=1,i_2=0,i_3=0),(i_1=1,i_2=0,i_3=1),(i_1=1,i_2=1,i_3=0),(i_1=1,i_2=1,i_3=1)\}$ and the C-volume is $\Delta_{(\mathbf{a},\mathbf{b}]}C= C(b_1,b_2,b_3)- C(b_1,b_2,a_3)- C(b_1,a_2,b_3)+ C(b_1,a_2,a_3)- C(a_1,b_2,b_3)+ C(a_1,b_2,a_3)+ C(a_1,a_2,b_3)- C(a_1,a_2,a_3)$

Give formulas for discrete and mixed margins likelihoods from Joe

Mention extension to individual margins with both discrete and continuous components [@gunawan_mixed_2018]

# R Session Information

```{r, echo=FALSE}
sessionInfo()
```

<!-- Code appendix if knitting to pdf -->
`r if (!knitr::is_html_output()) '# Code Appendix'` 

```{r, ref.label='ex1-a', echo=!knitr::is_html_output(), eval=FALSE}
```

```{r, ref.label='ex1-b', echo=!knitr::is_html_output(), eval=FALSE}
```

```{r, ref.label='norm-cop-ex', echo=!knitr::is_html_output(), eval=FALSE}
```

```{r, ref.label='tail-dep-ex', echo=!knitr::is_html_output(), eval=FALSE}
```

```{r, ref.label='br-a', echo=!knitr::is_html_output(), eval=FALSE}
```

